PGxMine: Text mining for curation of PharmGKB

### 2019-09-30 paper/paper.Rmd
Update a paper reference
Newer version of paper, lots of edits.

---
title: "PGxMine: Text mining for curation of PharmGKB"
#output: 
#  bookdown::pdf_document2
#  html_document: 
#    fig_caption: yes
#params:
#   region: east
#editor_options: 
#  chunk_output_type: console
site: bookdown::bookdown_site
documentclass: book
output:
#  bookdown::html_document2
  bookdown::word_document2
#  bookdown::pdf_document2
bibliography: pgxmine.bibtex
csl: world-neurosurgery.csl
---

@@

```{r pgxmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('figures.R')
source('pharmgkb_comparison.R')
invisible(dev.off())
```

@@

Precision medicine tailors treatment to individuals personal data including differences in their genome. The Pharmacogenomics Knowledgebase (PharmGKB) provides highly curated information on the effect of genetic variation on drug response and side effects for a wide range of drugs. PharmGKB’s scientific curators triage, review and annotate a large number of papers each year but the task is challenging. We present the PGxMine resource, a text-mined resource of pharmacogenomic associations from all accessible published literature to assist in curation of PharmGKB. We developed a supervised machine learning pipeline to extract associations between a variant (DNA and protein changes, star alleles and dbSNP identifiers) and a chemical. PGxMine covers `r paper.chemicalCount` chemicals and `r paper.variantCount` variants and contains `r paper.mentionCount` mentions of pharmacogenomic associations across `r paper.pmidCount` papers. An evaluation by PharmGKB curators found that 57 of the top 100 associations not found in PharmGKB led to 83 curatable papers and a further 24 associations would likely lead to curatable papers through citations. The results can be viewed at https://pgxmine.pharmgkb.org/ and data can be downloaded at LINK.

@@

Precision medicine has been described as getting patients "the right drug at the right dose at the right time" [@hamburg2010path]. It has been shown that genetics can play a large part in whether the drug and dose are right for that particular patient and numerous diagnostic tests have been developed to test for genetic variants related to drug efficacy [@relling2015pharmacogenomics]. These pharmacogenomic variants encompass germline variants or somatic variants in cancer. They may affect the direct target of a drug, potentially by disrupting binding pockets, (e.g. T790M for many EGFR inhibitors [@zhou2009novel]), the metabolizing enzymes that may affect drug concentrations, by affecting enzyme efficiency (e.g. the cytochrome P450 family of enzymes [@tornio2018cytochrome]), or drug transporters that may affect the disposition of the drugs and influence efficacy and toxicity.  Response to several drugs is now known to be affected greatly by individual genetic diversity. For example, the dosing of the blood-thinner warfarin is known to be influenced by variation in CYP2C9 and VKORC1 genes [@johnson2011clinical].

Cataloging this expanding knowledge of pharmacogenomic variants is the goal of the PharmGKB. Launched in 2000 [@whirl2012pharmacogenomics], it collects, curates and disseminates knowledge about gene-drug associations, many of which are clinically actionable,   and provides resources for both researchers and clinicians. A group of expert curators searches the scientific literature for relevant pharmacogenomic papers and add their findings to the knowledge base. However, managing the curation of a biomedical knowledge base is a very challenging task. As precision medicine efforts increase and the cost of sequencing decreases, more genetic biomarkers will be found that might prove to be clinically actionable. It is already a huge challenge to identify relevant papers and these technological developments will inevitably lead to an increase in the number of published papers and an increased burden on the PharmGKB curators.

Machine learning methods are used to assist in biomedical knowledge base curation in two different ways. A document-level approach attempts to identify documents that would be appropriate for curation into the knowledge base using word frequencies, metadata, and other factors. This approach was taken for the ORegAnno database [@aerts2008text], Gene Expression Database [@jiang2019effective] and for mouse genomics [@dowell2009integrating]. A mention-level approach attempts to extract the main associations that would be curated in the knowledge base and aggregating them to identify missing knowledge linked to curatable papers. The CIViCmine approach extracted mentions of clinically relevant cancer variants to identify the most frequently discussed that should be curated into the CIViC database [@lever2018text]. This method is linked with automated knowledge base construction methods which have been used in part to construct the STRING database of protein interactions [@szklarczyk2014string] and the CancerMine database of cancer genes [@lever2019cancermine] directly from published literature.

In order to extract pharmacogenomic variants associated with specific chemicals, we propose to use a biomedical relation extraction method. The field of biomedical relation extraction focuses on extracting mentions of particular relations between entities (e.g. genes, drugs, variants, etc) in unstructured text. Dependency-parse based methods have seen great success in this area [@lever2017painless] while deep learning methods achieve similar performance but face issues with overfitting [@mehryary2016deep]. Relation extraction relies on first identifying mentions of specific entities, e.g. chemicals and variants, using entity extraction tools such as BANNER [@leaman2008banner] and tmVar [@wei2017tmvar]. The Pubtator resource has provided PubMed abstracts annotated with chemicals, diseases, genes, species, and variants [@wei2013pubtator]. Numerous projects have built on top of the annotations provided by PubTator [@percha2018global,@mahmood2016dimex]. Recently, the Pubtator Central (PTC) resource expanded this to open-access Pubmed Central full-text articles [@wei2019pubtator].

In this paper, we present PGxMine, a text-mined resource of pharmacogenomic associations between chemicals and variants. We build upon the PubTator Central resource and use a text-alignment method to extract specific mentions of chemicals, genes, and variants in the text. We further identify star alleles that are particularly important/unique and frequently appears in pharmacogenomic literature (e.g. CYP2D6\*2) in text. We parse and extract sentences that mention a chemical and variant and filter to enrich for sentences that likely discuss pharmacogenomic associations. We build classifiers to extract pharmacogenomic associations and apply these to all sentences accessible from PubMed abstracts and PubMed Central full-text papers. These results are then presented to the PharmGKB team to assist in their curation efforts.

@@

(ref:pipeline) An overview of the full PGxMine system. The input data sources on the left (PubMed, etc) and PubTator Central are combined through a text alignment process to identify mentions of specific biological entities in published literature. Star Alleles (e.g. CYP2D6\*2) are then found using the gene annotations. Sentences are then filtered using keywords to enrich for pharmacogenomic topics. A Kindred supervised classifier is then trained and applied to identify specific variant/chemical associations. These are then filtered for high probability matches and collated to produce the three output files on the right.

@@

We downloaded the 26th June 2019 release of the Pubtator Central data which contains PubMed IDs associated with text mentions of chemicals, variants, genes, and other entities. The chemicals are mapped to MeSH terms, the genes are mapped to Entrez Gene IDs and where possible, the variants are mapped to dbSNP IDs (also known as rs IDs). We used the PubRunner infrastructure to download abstracts from PubMed and full-text papers from the PubMed Open Access (PMCOA) subset and PubMed Central Author Manuscript Collection (PMCAMC) and convert to BioC format [@comeau2013bioc].

@@

PubTator Central uses the tmVar 2 tool for extracting variant mentions. It can extract DNA-level mutations (e.g. c.93G>A), protein-level mutations (e.g. T790M) and dbSNP IDs (e.g. rs12979860). Where possible, tmVar 2 will map the DNA/protein variants back to a dbSNP ID using co-mentions of genes in the document to filter down candidate matches. Hence a mention of EGFR alongside T790M provides enough context to map it to rs121434569. We are primarily interested in those that can be mapped back to dbSNP IDs but kept all variant mentions as long as they contained a specific coordinate (i.e. filtering out cases such as c.G>A). For variants that are successfully mapped to a dbSNP ID by tmVar 2, PTC does not provide the normalized text form of the variant. For example, ‘an EGFR threonine-to-methionine mutation at residue 790’ is mapped successfully to rs121434569 but we also need the HGVS normalized form, p.T790M. We implemented a re-normalization stage using regular expressions to capture the variant types output by tmVar 2.

@@

We extract all sentences that mention at least one variant (including star alleles) and one chemical from our filtered list. In an initial trial of this system focussing only on mentions of specific dbSNP IDs, we found that sentences were highly likely to mention pharmacogenomic events. However, when all variants are included, sentences are much less likely to discuss a pharmacogenomic association. Any machine learning system would then have a large challenge with such an imbalanced problem. We, therefore, filter with a customized set of keywords (in Supplementary Materials) that increase the likelihood of a pharmacogenomic association or for a mention of a specific dbSNP ID.

@@

The previous steps provide a large set of co-occurrences between variants and chemicals. But many of them are not describing a pharmacogenomic association. For instance, some of the sentences are very long and the chemical and variant are discussed in different contexts. Therefore we build relation classifiers to identify the pharmacogenomic relations.

Due to a class balance difference, we split the dataset into two. The first set is all potential chemical/variant where the variant is a specific RS ID or a star allele. The second set is all remaining chemical/variant associations which contain variants for specific DNA or protein changes. We worked on the two sets separately because, empirically, the first set has a higher likelihood for discussing pharmacogenomic associations. For each set, we annotated 500 sentences for pharmacogenomic associations. Chemical/variant associations were annotated as pharmacogenomic if the variant is discussed as physically interacting with the chemical, as affecting the metabolism, concentration or resistance of the chemical, as causing adverse events related to the chemical, as related to addiction to the chemical, or is part of a clinical test for that chemical use. The PharmGKB curators are also interested in variants found to be negatively associated pharmacogenomically with a chemical, so we also annotated negative discoveries as positive sentences in our data set. In total, we had 1000 annotated sentences which for evaluation are split into 80%/20% training and test sets.

@@

We mapped the PGxMine set of associations to those found in PharmGKB. PharmGKB already contains some mappings directly to MeSH IDs but these needed to be supplemented with additional mappings from other sources. DrugBank provides a mapping from the MeSH ID used by PubTator Central to PharmGKB chemical IDs. Variants that have been mapped to dbSNPs or are star alleles can be mapped directly into PharmGKB. PharmGKB contains compound associations where multiple chemicals are associated with a variant. For simplicity, we unfold these associations so that all associations are from one chemical to one variant, allowing matching against the PGxMine data. For this comparison, we remove suballeles from star alleles (e.g. CYP3A5\*3A -> CYP3A5\*3). We also build a mapping for star alleles that map directly to a single RS ID to check if either is found in PharmGKB (e.g. POR\*28 maps to rs1057868). Further details of this mapping are explained in the Supplementary Materials. For each PGxMine association, we then check if the variant has been seen in PharmGKB, the chemical has been seen in PharmGKB and whether this particular association has been seen.

@@

The filtered and collated versions are then viewed through a viewer built using the R Shiny framework. We host a publicly accessible version at URL. This viewer (shown in Figure \@ref(fig:viewer)) shows the collated chemical-variant associations sorted in descending order by the number of papers that mention them.  These results can then be filtered by a chemical, variant or gene, whether this data exists in PharmGKB and the type of variant, e.g. star allele. There are also Download buttons to facilitate downloading only a subset of the data. When a row in the top table is selected, a lower table is populated with all the sentences that discuss that particular association. Metadata is provided for the source of the sentence with links to the original paper.

@@

Using an 80%/20% training and test split, we evaluate the two classifiers independently. Figure \@ref(fig:prcurves) shows precision-recall curve plots of the three classifiers. The black line shows the selected high threshold (0.75) that provides higher precision (`r paper.precision1`% and `r paper.precision2`% for the two classifiers) with the tradeoff of lower recall (`r paper.recall1`% and `r paper.recall2`%).

@@

By applying the two classifiers we identify `r paper.mentionCount` mentions of pharmacogenomic associations in `r paper.sentenceCount` sentences across `r paper.pmidCount` papers. `r paper.titleOrAbstract`% of these associations are in the title or abstract. When collated, we find a total of `r paper.associationCount` unique gene/variant associations. Of all the variant mentions, `r paper.normalizedToRSID`% can be normalized to a rsID and `r paper.starAlleles`% are star alleles.

As would be expected, there is a long-tail of associations with `r paper.singlePapers`% of the unique associations being discussed in only a single paper. These singletons are of interest to the PharmGKB curators as the aim to improve coverage for rarer and less frequently discussed associations. We also find that the associations are found across all subsections of papers, even including the author contributions section.

@@

We compare the associations extracted from the literature with the contents of the PharmGKB database. Figure \@ref(fig:comparison) shows the overlap of associations and specific papers. The scale of the PharmGKB database cannot be understated and this figure illustrates the vast number of associations that have already been manually curated. `r paper.pharmGKBPercentageOverlap`% of associations found in PGxMine have previously been curated into PharmGKB but `r paper.associationsNotInPharmGKB` could be curatable associations. As we identify mentions of associations, many of which will be in papers discussing previous work and not a new result, we wouldn't expect to see a large overlap in the articles in PGxMine and PharmGKB. Nevertheless, we do see a sizeable overlap of papers based on PubMed IDs suggesting that PGxMine does often directly identify the appropriate paper for curation.

@@

Our results suggest that PGxmine will assist PharmGKB curators but there are several notable limitations. The results of PGxMine rely entirely on the results of PubTator Central. We have explored running the individual NER tools on the large corpus ourselves but the computational cost, in terms of time and memory usage, is immense. Updates to our resource will, therefore, depend on the update frequency of PubTator Central which we hope to be frequent. This reliance of PGxMine on PubTator Central also means that we inherit any limitations of the NER tools used. While the NER tools are state-of-the-art for the entities that they are extracting, mistakes will still be made. We attempted to clean up issues by requiring longer terms and filtering out certain drug categories.

PGxMine is also limited to associations found within the same sentence. This is a problem that all current information extraction systems face as the false positive rate explodes for systems that attempt to extract relations across sentences [@lever2016verse]. This is a ripe area for further research and an annotated coreference dataset may push forward performance in this area in the future.

The tradeoff of recall and precision is a challenging one for many machine learning problems. Curators become frustrated with a system that has too many false positives and are understanding of a system that does not have 100% coverage of associations. We, therefore, target high-precision low-recall for this problem, especially when the frequently discussed associations are found in hundreds of papers, meaning that they will likely be discovered by our pipeline. For example, the most frequently extracted association is `r paper.topAssoc_chemical` with `r paper.topAssoc_variant` in `r paper.topAssoc_paperCount` papers. This means we should capture frequently discussed associations that aren't yet curated into PharmGKB. We find that the vast majority of extracted associations (`r paper.singlePapers`%) are only mentioned in one paper. While more frequently discussed associations are of greater interest to PharmGKB, these singletons provide a valuable trove for further curation.

@@

We have presented PGxMine, a knowledge base of pharmacogenomic associations to support curation of PharmGKB database. We developed a method for extracting knowledge from the new PubTator Central resource using a text alignment system coupled with a method for identifying star alleles and proven relation extraction methods. Our results can be updated regularly as PubTator Central is updated. We make the code and supplementary materials available with an MIT license (https://github.com/jakelever/pgxmine) and data available under CC0 license (TODO: INSERT Zenodo URL) with the understanding that all data is extracted automatically and has not been vetted by the PharmGKB curation team.

>>>>

---
title: "PGxMine: Text mining for curation of PharmGKB"
#output: 
#  bookdown::pdf_document2
#  html_document: 
#    fig_caption: yes
#params:
#   region: east
#editor_options: 
#  chunk_output_type: console
site: bookdown::bookdown_site
documentclass: book
#output:
#  bookdown::html_document2
#  bookdown::word_document2
#  bookdown::pdf_document2
output:
   pdf_document:
      latex_engine: xelatex
bibliography: pgxmine.bib
csl: world-neurosurgery.csl
---

@@

```{r pgxmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('figures.R')
source('pharmgkb_comparison.R')
source('subsections.R')
invisible(dev.off())
```

@@

Precision medicine tailors treatment to individuals personal data including differences in their genome. The Pharmacogenomics Knowledgebase (PharmGKB) provides highly curated information on the effect of genetic variation on drug response and side effects for a wide range of drugs. PharmGKB’s scientific curators triage, review and annotate a large number of papers each year but the task is challenging. We present the PGxMine resource, a text-mined resource of pharmacogenomic associations from all accessible published literature to assist in the curation of PharmGKB. We developed a supervised machine learning pipeline to extract associations between a variant (DNA and protein changes, star alleles and dbSNP identifiers) and a chemical. PGxMine covers `r paper.chemicalCount` chemicals and `r paper.variantCount` variants and contains `r paper.mentionCount` mentions of pharmacogenomic associations across `r paper.pmidCount` papers. An evaluation by PharmGKB curators found that 57 of the top 100 associations not found in PharmGKB led to 83 curatable papers and a further 24 associations would likely lead to curatable papers through citations. The results can be viewed at https://pgxmine.pharmgkb.org/ and code can be downloaded at https://github.com/jakelever/pgxmine.

@@

Precision medicine has been described as getting patients "the right drug at the right dose at the right time" [@hamburg2010path]. It has been shown that genetics can play a significant part in whether the drug and dose are right for that particular patient and numerous diagnostic tests have been developed to test for genetic variants related to drug efficacy [@relling2015pharmacogenomics]. These pharmacogenomic variants encompass germline variants or somatic variants in cancer. They may affect the direct target of a drug, potentially by disrupting binding pockets, (e.g. T790M for many EGFR inhibitors [@zhou2009novel]), the metabolizing enzymes that may affect drug concentrations, by affecting enzyme efficiency (e.g. the cytochrome P450 family of enzymes [@tornio2018cytochrome]), or drug transporters that may affect the disposition of the drugs and influence efficacy and toxicity.  Response to several drugs is now known to be affected greatly by individual genetic diversity. For example, the dosing of the blood-thinner warfarin is known to be influenced by variation in CYP2C9, VKORC1, and CYP4F2 genes [@johnson2017clinical].

Cataloging this expanding knowledge of pharmacogenomic variants is the goal of the PharmGKB. Launched in 2000 [@whirl2012pharmacogenomics], it collects, curates and disseminates knowledge about gene-drug associations, many of which are clinically actionable, and provides resources for both researchers and clinicians. A group of expert curators searches the scientific literature for relevant pharmacogenomic papers and add their findings to the knowledge base. However, managing the curation of a biomedical knowledge base is a very time-consuming and challenging task. As precision medicine efforts increase and the cost of sequencing decreases, more genetic biomarkers will be found that might prove to be clinically actionable. It is already a huge challenge to identify relevant papers and these technological developments will inevitably lead to an increase in the number of published papers and an increased burden on the PharmGKB curators.

Machine learning methods are used to assist in biomedical knowledge base curation in two different ways. A document-level approach attempts to identify documents that would be appropriate for curation into the knowledge base using word frequencies, metadata, and other factors. This approach was taken for the ORegAnno database [@aerts2008text], Gene Expression Database [@jiang2019effective] and mouse genomics [@dowell2009integrating]. A mention-level approach attempts to extract the main associations that would be curated in the knowledge base and aggregating them to identify missing knowledge linked to curatable papers. The CIViCmine approach extracted mentions of clinically relevant cancer variants to identify the most frequently discussed that should be curated into the CIViC database [@lever2018text]. This method is linked with automated knowledge base construction methods which have been used in part to construct the STRING database of protein interactions [@szklarczyk2014string] and the CancerMine database of cancer genes [@lever2019cancermine] directly from published literature.

To extract pharmacogenomic variants associated with specific chemicals, we propose to use a biomedical relation extraction method. The field of biomedical relation extraction focuses on extracting mentions of particular relations between entities (e.g. genes, drugs, variants, etc) in unstructured text. Dependency-parse based methods have seen great success in this area [@lever2017painless] while deep learning methods achieve similar performance but face issues with overfitting [@mehryary2016deep]. Relation extraction relies on first identifying mentions of specific entities, e.g. chemicals and variants, using entity extraction tools such as BANNER [@leaman2008banner] and tmVar [@wei2017tmvar]. The Pubtator resource has provided PubMed abstracts annotated with chemicals, diseases, genes, species, and variants [@wei2013pubtator]. Numerous projects have built on top of the annotations provided by PubTator [@percha2018global,@mahmood2016dimex]. Recently, the Pubtator Central (PTC) resource expanded this to open-access Pubmed Central full-text articles [@wei2019pubtator].

In this paper, we present PGxMine, a text-mined resource of pharmacogenomic associations between chemicals and variants. We build upon the PubTator Central resource and use a text-alignment method to extract specific mentions of chemicals, genes, and variants in the text. We further identify star alleles that are particularly important/unique and frequently appears in pharmacogenomic literature (e.g. CYP2D6\*2) in text. We parse and extract sentences that mention a chemical and variant and filter to enrich for sentences that likely discuss pharmacogenomic associations. We build classifiers to extract pharmacogenomic associations and apply these to all sentences accessible from PubMed abstracts and PubMed Central full-text papers. These results are then presented to the PharmGKB scientific curators to assist in their curation efforts.

@@

(ref:pipeline) An overview of the full PGxMine system. The input data sources on the left (PubMed, etc) and PubTator Central are combined through a text alignment process to identify mentions of specific biological entities in published literature. Star Alleles (e.g. CYP2D6\*2) are then found using gene annotations. Sentences are filtered using keywords to enrich for pharmacogenomic topics. A Kindred supervised classifier is then trained and applied to identify specific variant/chemical associations. These are then filtered for high probability matches and collated to produce the three output files on the right.

@@

We downloaded the 21st August 2019 release of the Pubtator Central data which contains PubMed IDs associated with text mentions of chemicals, variants, genes, and other entities. It uses TaggerOne [@leaman2016taggerone] for chemicals, tmVar 2.0 [@wei2017tmvar] for variants and GNormPlus [@wei2015gnormplus] for genes. It provides mappings for chemical mentions to MeSH terms, genes to Entrez Gene IDs and where possible, variants to dbSNP IDs [@sherry2001dbsnp] (also known as rs IDs). We used the PubRunner infrastructure to download abstracts from PubMed and full-text papers from the PubMed Open Access (PMCOA) subset and PubMed Central Author Manuscript Collection (PMCAMC) and convert to BioC format [@comeau2013bioc].

@@

PubTator Central uses the tmVar 2 tool for extracting variant mentions [@wei2017tmvar]. It can extract DNA-level mutations (e.g. c.93G>A), protein-level mutations (e.g. T790M) and dbSNP IDs (e.g. rs12979860). Where possible, tmVar 2 will map the DNA/protein variants back to a dbSNP ID using co-mentions of genes in the document to filter down candidate matches. Hence a mention of EGFR alongside T790M provides enough context to map it to rs121434569. We are primarily interested in those that can be mapped back to dbSNP IDs but kept all variant mentions as long as they contained a specific coordinate (i.e. filtering out cases such as c.G>A). For variants that are successfully mapped to a dbSNP ID by tmVar 2, PTC does not provide the normalized text form of the variant. For example, ‘an EGFR threonine-to-methionine mutation at residue 790’ is mapped successfully to rs121434569 but we also need the HGVS normalized form, p.T790M. We implemented a re-normalization stage using regular expressions to capture the variant types output by tmVar 2.

@@

We extract all sentences that mention at least one variant (including star alleles) and one chemical from our filtered list. In an initial trial of this system focussing only on mentions of specific dbSNP IDs, we found that sentences were highly likely to mention pharmacogenomic events. However, when all variants were included, sentences were much less likely to discuss a pharmacogenomic association. Any machine learning system would then have a large challenge with such an imbalanced problem. We, therefore, filtered with a customized set of keywords (in Supplementary Materials) that increased the likelihood of a pharmacogenomic association or for a mention of a specific dbSNP ID.

@@

The previous steps provide a large set of co-occurrences between variants and chemicals. But many of them are not describing a pharmacogenomic association. For instance, some of the sentences are very long and the chemical and variant are discussed in different contexts. Therefore we built relation classifiers to identify the pharmacogenomic relations.

Due to a class balance difference, we split the dataset into two. The first set is all potential chemical/variant where the variant is a specific rs ID or a star allele. The second set is all remaining chemical/variant associations which contain variants for specific DNA or protein changes. We worked on the two sets separately because, empirically, the first set has a higher likelihood for discussing pharmacogenomic associations. For each set, we annotated 500 sentences for pharmacogenomic associations. Chemical/variant associations were annotated as pharmacogenomic if the variant was discussed as physically interacting with the chemical, as affecting the metabolism, concentration or resistance of the chemical, as causing adverse events related to the chemical, as related to addiction to the chemical, or as part of a clinical test for that chemical use. The PharmGKB curators were also interested in variants found to be negatively associated pharmacogenomically with a chemical. Thus, we also annotated negative discoveries as positive sentences in our data set. In total, we had 1000 annotated sentences which for evaluation are split into 80%/20% training and test sets. We include statistics on the annotated dataset in the supplementary materials.

@@

We mapped the PGxMine set of associations to those found in PharmGKB. PharmGKB already contains some mappings directly to MeSH IDs but these needed to be supplemented with additional mappings from other sources. DrugBank provides a mapping from the MeSH ID used by PubTator Central to PharmGKB chemical IDs. Variants that have been mapped to dbSNPs or are star alleles can be mapped directly into PharmGKB. PharmGKB contains compound associations where multiple chemicals are associated with a variant. For simplicity, we unfold these associations so that all associations are from one chemical to one variant, allowing matching against the PGxMine data. For this comparison, we remove suballeles from star alleles (e.g. CYP3A5\*3A -> CYP3A5\*3). We also built a mapping for star alleles that map directly to a single rs ID to check if either was found in PharmGKB (e.g. POR\*28 maps to rs1057868). Further details of this mapping are explained in the Supplementary Materials. For each PGxMine association, we then check if the variant has been seen in PharmGKB, the chemical has been seen in PharmGKB and whether this particular association has been seen.

@@

The filtered and collated versions are then viewed through a viewer built using the R Shiny framework. This viewer (shown in Figure \@ref(fig:viewer)) shows the collated chemical-variant associations sorted in descending order by the number of papers that mention them.  These results can then be filtered by a chemical, variant or gene, whether this data exists in PharmGKB and the type of variant, e.g. star allele. There are also Download buttons to facilitate downloading only a subset of the data. When a row in the top table is selected, a lower table is populated with all the sentences that discuss that particular association. These sentences, which may be from the same paper or across many papers, are ordered by publication date and additional metadata is provided for the source of the sentence with links to the original paper.

@@

Using an 80%/20% training and test split, we evaluated the two classifiers independently. Figure \@ref(fig:prcurves) shows precision-recall curve plots of the three classifiers. The black line shows the selected high threshold (0.75) that provides higher precision (`r paper.precision1`% and `r paper.precision2`% for the two classifiers) with the tradeoff of lower recall (`r paper.recall1`% and `r paper.recall2`%).

@@

By applying the two classifiers we identify `r paper.mentionCount` mentions of pharmacogenomic associations in `r paper.sentenceCount` sentences across `r paper.pmidCount` papers. `r paper.titleOrAbstract`% of these associations were in the title or abstract. When collated, we find a total of `r paper.associationCount` unique gene/variant associations. Of all the variant mentions, `r paper.normalizedToRSID`% can be normalized to an rs ID and `r paper.starAlleles`% are star alleles.

```{r subsectionPlot, echo=F, eval=T, fig.width=10, fig.asp=.5,  fig.cap='(ref:subsectionPlot)'}
grid.arrange(subsectionPlot)
```

(ref:subsectionPlot) From the full-text papers accessed through PubMed Central, associations are extracted from all sections of the full-text paper.

As would be expected, there is a long-tail of associations with `r paper.singlePapers`% of the unique associations being discussed in only a single paper. These singletons are of interest to the PharmGKB curators as the aim to improve coverage for rarer and less frequently discussed associations. We also find that the associations were found across all subsections of papers (Figure \@ref(fig:subsectionPlot)).

@@

We compared the associations extracted from the literature with the contents of the PharmGKB knowledgebase downloaded on `r paper.pharmgkbDate`. Figure \@ref(fig:comparison) shows the overlap of associations and specific papers. The scale of the PharmGKB knowledgebase cannot be understated and this figure illustrates the vast number of associations that have already been manually curated. `r paper.pharmGKBPercentageOverlap`% of associations found in PGxMine have previously been curated into PharmGKB but `r paper.associationsNotInPharmGKB` could be curatable associations. As we identify mentions of associations, many of which will be in papers discussing previous work and not a new result, we wouldn't expect to see a large overlap in the articles in PGxMine and PharmGKB. Nevertheless, we do see a sizeable overlap of papers based on PubMed IDs suggesting that PGxMine does often directly identify the appropriate paper for curation.

@@

Our results suggest that PGxmine will assist PharmGKB curators but there are several notable limitations. The results of PGxMine rely entirely on the results of PubTator Central. We have explored running the individual NER tools on the large corpus ourselves but the computational cost, in terms of time and memory usage, is immense. Updates to our resource will, therefore, depend on the update frequency of PubTator Central which we hope to be frequent. This reliance on PGxMine on PubTator Central also means that we inherit any limitations of the NER tools used. While the NER tools are state-of-the-art for the entities that they are extracting, mistakes will still be made. We attempted to clean up issues by requiring longer terms and filtering out certain drug categories.

However, the main limiting factor is the poor access to full-text papers for text mining. `r paper.mentionsInFullTextPerc`% (`r paper.mentionCount`/`r paper.mentionsInFullText`) mentions of pharmacogenomic associations were extracted from the main text of the article. Only a fraction of the biomedical literature is accessible for text mining. With only `r paper.pharmgkbPaperHasFullTextPerc`% (`r paper.pharmgkbPaperHasFullText`/`r paper.pharmgkbPaperNum`) of the papers curated in PharmGKB having full-text available, it is understandable that PGxMine fails to catch many of the previously curated papers. PGxMine is also limited to associations found within the same sentence. This is a problem that all current information extraction systems face as the false positive rate explodes for systems that attempt to extract relations across sentences [@lever2016verse]. This is a ripe area for further research and an annotated coreference dataset may push forward performance in this area in the future.

The tradeoff of recall and precision is a challenging one for many machine learning problems. Curators become frustrated with a system that has too many false positives and are understanding of a system that does not have 100% coverage of associations. We, therefore, target high-precision low-recall for this problem, especially when the frequently discussed associations are found in hundreds of papers, meaning that they will likely be discovered by our pipeline. For example, the most frequently extracted association is `r paper.topAssoc_chemical` with `r paper.topAssoc_variant` in `r paper.topAssoc_paperCount` papers. This means we should capture frequently discussed associations that aren't yet curated into PharmGKB. We find that the vast majority of extracted associations (`r paper.singlePapers`%) are only mentioned in one paper. While more frequently discussed associations are of greater interest to PharmGKB, these singletons provide a valuable trove for further curation. The most common mistake that the classifier makes is incorrectly matching a drug with a variant in a different clause of a sentence. This occurs most frequently in sentences with multiple drugs and variants.

@@

We have presented PGxMine, a knowledge base of pharmacogenomic associations to support curation of PharmGKB database. We developed a method for extracting knowledge from the new PubTator Central resource using a text alignment system coupled with a method for identifying star alleles and proven relation extraction methods. Our results can be updated regularly as PubTator Central is updated. We make the code and supplementary materials available with an MIT license (https://github.com/jakelever/pgxmine) and data available under CC0 license (https://doi.org/10.5281/zenodo.3360930) with the understanding that all data is extracted automatically and has not been vetted by the PharmGKB curation team.

## Acknowledgements

This work is supported by the NIH/NIGMS grant GM61374 (TEK, RBA), NIH/NLM grant LM005652 (RBA) and NIH/NCATS grant TR002515 (RBA).
